---
title: "FairHire AI: Statutory Bias Audit Report"
author: "Group 4: Risk Ready"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: true
---
Bias testing conducted by Group 4 on the data set from Kaggle <br>
Data Source & Specification <br>
Dataset Name: Resume Screening Dataset (200k Candidates)<br>
Provider: Rhythm Ghai via Kaggle <br>
Format: Tabular (CSV) <br>
Sample Size: 200,000 synthetic candidate profiles <br>
Key Variables: Education Level, Age, University Tier, Years of Experience, and Hiring Outcome (Target). <br>
Used in the project as a primary testing ground for the FairHire AI bias audit. It is utilized to simulate high-volume recruitment scenarios and calculate Impact Ratios across protected demographic groups.<br>

```{r setup, include=FALSE}
# CHUNK 1: ENVIRONMENT SETUP
# include=FALSE hides installation messages from the report

# Auto-install missing packages
packages <- c("reticulate", "tidyverse", "data.table", 
               "janitor", "scales", "knitr", "kableExtra")
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if (length(new_packages)) install.packages(new_packages)

library(reticulate)
library(tidyverse)
library(data.table)
library(janitor)
library(scales)
library(knitr)
library(kableExtra)

# FIX: Set knitr options for clean output throughout the report
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r data_processing, include=FALSE}
# CHUNK 2: DATA ACQUISITION & CLEANING
# include=FALSE runs this silently — variables are available to all chunks below

# 1. Download via kagglehub
kh   <- import("kagglehub")
path <- kh$dataset_download("rhythmghai/resume-screening-dataset-200k-candidates")

# 2. Find and load CSV
all_files <- list.files(path, recursive = TRUE, full.names = TRUE)
csv_file  <- all_files[str_detect(all_files, "\\.csv$")][1]   # FIX: escaped dot in regex

if (is.na(csv_file)) stop("ERROR: No CSV file found in downloaded dataset path.")

raw_data <- fread(csv_file)

# 3. Clean and engineer features
audit_prep <- raw_data %>%
  as_tibble() %>%
  clean_names() %>%
  mutate(
    # EU AI Act: age grouping to test for ageism
    age_group = case_when(
      age < 25             ~ "Entry / Gen Z",
      age >= 25 & age < 40 ~ "Mid-Career / Millennial",
      age >= 40            ~ "Senior / Gen X+",
      TRUE                 ~ "Unknown"          # FIX: catch-all for NAs
    ),
    # Proxy bias: university tier flag
    is_elite_school = ifelse(university_tier == "Tier 1", "Yes", "No"),
    # FIX: ensure 'hired' is numeric for sum() to work correctly
    hired = as.integer(hired)
  )

# 4. Education-level bias audit
edu_audit <- audit_prep %>%
  group_by(education_level) %>%
  summarise(
    total          = n(),
    selected       = sum(hired, na.rm = TRUE),
    selection_rate = selected / total,
    .groups        = "drop"
  ) %>%
  mutate(
    impact_ratio = selection_rate / max(selection_rate),
    # FIX: status labels must exactly match scale_fill_manual values below
    status = ifelse(impact_ratio < 0.8, "FAIL (Bias)", "PASS")
  )
```
<br>
NYC 144 Law Impact Ratio must be calculated by comparing the selection rate of each group with the highest rate. 
<br>
Treshold set at 0.8.
<br>
```{r table, echo=FALSE}
# CHUNK 3: EDUCATION AUDIT TABLE
# echo=FALSE shows the table but hides the code

kable(
  edu_audit %>%                              # FIX: format columns for readability
    mutate(
      selection_rate = percent(selection_rate, accuracy = 0.1),
      impact_ratio   = round(impact_ratio, 3)
    ),
  col.names = c("Education Level", "Total", "Selected", 
                "Selection Rate", "Impact Ratio", "Status"),
  caption   = "Table 1: Education Level Selection Rates (NYC LL 144 / 4/5ths Rule)"
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) %>%
  row_spec(
    which(edu_audit$status == "FAIL (Bias)"),
    background = "#ffe0e0"                   # highlight failing rows in red
  )
```

```{r edu_chart, echo=FALSE, fig.width=9, fig.height=5}
# CHUNK 4: EDUCATION IMPACT RATIO CHART

ggplot(edu_audit, aes(
    x    = reorder(education_level, -impact_ratio),
    y    = impact_ratio,
    fill = status
  )) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red", linewidth = 1.2) +
  geom_text(
    aes(label = round(impact_ratio, 2)),
    vjust    = -0.5,
    fontface = "bold",
    size     = 3.5
  ) +
  # FIX: labels now exactly match the 'status' values set in data_processing
  scale_fill_manual(
    values = c("PASS" = "#55efc4", "FAIL (Bias)" = "#ff7675"),
    labels = c("PASS" = "✅ PASS",  "FAIL (Bias)" = "⚠️ FAIL (Bias)")
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1.15)
  ) +
  labs(
    title   = "Impact Ratio by Education Level (NYC LL 144 Compliance)",
    subtitle = "Red dashed line = 80% legal threshold for Disparate Impact",
    y       = "Impact Ratio",
    x       = "Education Level",
    fill    = "Compliance Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```
<br>
This chart visualizes the age disparity to monitor the age bias. 
<br>
```{r age_density, echo=FALSE, fig.width=9, fig.height=5}
# CHUNK 5: AGE SELECTION DENSITY CHART

ggplot(audit_prep, aes(x = age, fill = factor(hired))) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(
    values = c("0" = "#dfe6e9", "1" = "#0984e3"),
    labels = c("0" = "Rejected", "1" = "Hired")
  ) +
  labs(
    title    = "Selection Density by Age",
    subtitle = "Overlapping distributions suggest potential age-based disparate impact",
    x        = "Candidate Age",
    y        = "Density",
    fill     = "Outcome"
  ) +
  theme_minimal(base_size = 12)
```