---
title: "FairHire AI: NYC Local Law 144 Statutory Bias Audit Report"
author: "Group 4: Risk Ready"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
# ── CHUNK 1: GLOBAL SETUP ─────────────────────────────────────────────────────
# include=FALSE hides all setup noise from the final report

knitr::opts_chunk$set(
  echo    = FALSE,   # hide code in output
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5.5
)

# Auto-install missing packages
packages <- c("tidyverse", "janitor", "scales", "knitr", "kableExtra")
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if (length(new_packages)) install.packages(new_packages, quiet = TRUE)

library(tidyverse)
library(janitor)
library(scales)
library(knitr)
library(kableExtra)
```

```{r data_loading, include=FALSE}
# ── CHUNK 2: DATA LOADING ─────────────────────────────────────────────────────
# include=FALSE keeps data loading silent in the report

# setwd ensures relative paths work regardless of who runs this
setwd(dirname(rstudioapi::getSourceEditorContext()$path))

file_path <- "Data/FairHire_Master_Resumes.csv"

# Fallback to absolute path if relative path fails (e.g. when knitting)
if (!file.exists(file_path)) {
  file_path <- "C:/GMU_Work/ME-575-AI-Design-and-Deployment-Risks/Data/FairHire_Master_Resumes.csv"
}

if (!file.exists(file_path)) {
  stop(paste0(
    "File not found. Run file.choose() in the console to locate ",
    "FairHire_Master_Resumes.csv and update file_path."
  ))
}

data <- read_csv(file_path, show_col_types = FALSE) %>% clean_names()
```

```{r simulation, include=FALSE}
# ── CHUNK 3: DEMOGRAPHIC SIMULATION & SCORING ─────────────────────────────────
set.seed(42)

audit_data <- data %>%
  mutate(
    gender = sample(
      c("Male", "Female", "Non-Binary"),
      n(), replace = TRUE,
      prob = c(0.48, 0.48, 0.04)
    ),
    race_ethnicity = sample(
      c("White", "Black", "Hispanic", "Asian", "Other"),
      n(), replace = TRUE
    ),
    ai_selection = as.integer(
      sample(c(0, 1), n(), replace = TRUE, prob = c(0.7, 0.3))
    )
  )

# ── NYC LL144 IMPACT RATIO FUNCTION ───────────────────────────────────────────
calculate_impact_ratio <- function(df, group_var) {
  df %>%
    group_by({{ group_var }}) %>%
    summarise(
      total_candidates = n(),
      selected         = sum(ai_selection, na.rm = TRUE),
      selection_rate   = selected / total_candidates,
      .groups          = "drop"
    ) %>%
    mutate(
      max_rate     = max(selection_rate),
      impact_ratio = selection_rate / max_rate,
      status       = ifelse(impact_ratio < 0.8,
                            "FAIL (Potential Bias)", "PASS")
    ) %>%
    arrange(desc(impact_ratio))
}

gender_audit <- calculate_impact_ratio(audit_data, gender)
race_audit   <- calculate_impact_ratio(audit_data, race_ethnicity)
```

```{r debiasing, include=FALSE}
# ── CHUNK 4: REWEIGHING DEBIASING ─────────────────────────────────────────────
overall_rate <- mean(audit_data$ai_selection)

group_rates <- audit_data %>%
  group_by(race_ethnicity) %>%
  summarise(group_rate = mean(ai_selection), .groups = "drop")

audit_data_debiased <- audit_data %>%
  left_join(group_rates, by = "race_ethnicity") %>%
  mutate(
    weight = overall_rate / group_rate,
    debiased_selection = as.integer(
      rbinom(n(), 1, prob = pmin(ai_selection * weight, 1))
    )
  )

calculate_impact_ratio_col <- function(df, group_var, selection_col) {
  df %>%
    group_by({{ group_var }}) %>%
    summarise(
      total_candidates = n(),
      selected         = sum({{ selection_col }}, na.rm = TRUE),
      selection_rate   = selected / total_candidates,
      .groups          = "drop"
    ) %>%
    mutate(
      max_rate     = max(selection_rate),
      impact_ratio = selection_rate / max_rate,
      status       = ifelse(impact_ratio < 0.8, "FAIL (Potential Bias)", "PASS")
    ) %>%
    arrange(desc(impact_ratio))
}

race_audit_before <- calculate_impact_ratio_col(
  audit_data_debiased, race_ethnicity, ai_selection) %>%
  mutate(model = "Before Debiasing")

race_audit_after <- calculate_impact_ratio_col(
  audit_data_debiased, race_ethnicity, debiased_selection) %>%
  mutate(model = "After Debiasing")

comparison <- bind_rows(race_audit_before, race_audit_after)
```

---

# Executive Summary

This report presents the results of a **statutory bias audit** conducted on the
FairHire AI automated resume screening system under **NYC Local Law 144 (2023)**,
which mandates annual third-party bias audits for all automated employment
decision tools (AEDTs) used in hiring within New York City.

The audit evaluates **Disparate Impact** across protected classes using the
**4/5ths Rule (80% Threshold)**: any demographic group whose selection rate
falls below 80% of the highest-selected group is flagged as a potential
compliance violation.

```{r summary_cards}
# ── AUDIT SNAPSHOT ────────────────────────────────────────────────────────────
total_candidates  <- nrow(audit_data)
total_selected    <- sum(audit_data$ai_selection)
overall_sel_rate  <- mean(audit_data$ai_selection)
gender_fails      <- sum(gender_audit$status == "FAIL (Potential Bias)")
race_fails        <- sum(race_audit$status   == "FAIL (Potential Bias)")

tibble(
  Metric = c(
    "Total Candidates Audited",
    "Total Selected by AI",
    "Overall Selection Rate",
    "Gender Groups Failing 80% Rule",
    "Race/Ethnicity Groups Failing 80% Rule"
  ),
  Value = c(
    format(total_candidates, big.mark = ","),
    format(total_selected,   big.mark = ","),
    percent(overall_sel_rate, accuracy = 0.1),
    paste0(gender_fails, " of ", nrow(gender_audit)),
    paste0(race_fails,  " of ", nrow(race_audit))
  )
) %>%
  kable(caption = "Table 1: Audit Snapshot") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) %>%
  row_spec(4:5,
    background = ifelse(c(gender_fails, race_fails) > 0, "#ffe0e0", "#e0ffe0")
  )
```

---

# Dataset Overview

The audit was conducted on resumes extracted from the FairHire AI resume
dataset. Demographic attributes (gender, race/ethnicity) and AI selection
outcomes were simulated to reflect realistic hiring distributions for
compliance testing purposes.

```{r dataset_table}
# ── DATASET PREVIEW ───────────────────────────────────────────────────────────
data %>%
  head(10) %>%
  kable(caption = "Table 2: Sample Resume Records (First 10 Rows)") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE
  ) %>%
  scroll_box(width = "100%")
```

---

# Gender Bias Audit

## Selection Rate Results {-}

```{r gender_table}
gender_audit %>%
  mutate(
    selection_rate = percent(selection_rate, accuracy = 0.1),
    impact_ratio   = round(impact_ratio, 4)
  ) %>%
  select(-max_rate) %>%
  rename(
    "Gender"           = gender,
    "Total Candidates" = total_candidates,
    "Selected"         = selected,
    "Selection Rate"   = selection_rate,
    "Impact Ratio"     = impact_ratio,
    "LL144 Status"     = status
  ) %>%
  kable(caption = "Table 3: Gender Selection Rates & Impact Ratios") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width        = FALSE
  ) %>%
  row_spec(
    which(gender_audit$status == "FAIL (Potential Bias)"),
    background = "#ffe0e0", bold = TRUE
  ) %>%
  row_spec(
    which(gender_audit$status == "PASS"),
    background = "#e8f8f0"
  )
```

## Impact Ratio Chart {-}

```{r gender_plot, fig.cap="Figure 1: Gender Impact Ratios — red dashed line marks the 80% legal threshold"}
ggplot(gender_audit, aes(
    x    = reorder(gender, -impact_ratio),
    y    = impact_ratio,
    fill = status
  )) +
  geom_col(width = 0.55, color = "white") +
  geom_hline(yintercept = 0.8, linetype = "dashed",
             color = "red", linewidth = 1) +
  annotate("text", x = 0.55, y = 0.84,
           label = "80% Legal Threshold",
           color = "red", size = 3.5, hjust = 0) +
  geom_text(
    aes(label = percent(impact_ratio, accuracy = 0.1)),
    vjust = -0.5, fontface = "bold", size = 4
  ) +
  scale_fill_manual(
    values = c("PASS"                  = "#A8E6CF",
               "FAIL (Potential Bias)" = "#FF8B94"),
    labels = c("PASS"                  = "✅ PASS",
               "FAIL (Potential Bias)" = "⚠️ FAIL")
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1.18)
  ) +
  labs(
    title    = "FairHire AI: Gender Impact Ratio Audit",
    subtitle = "NYC Local Law 144 Statutory Compliance Check",
    x        = "Gender Group",
    y        = "Impact Ratio",
    fill     = "Compliance Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    legend.position = "top"
  )
```

---

# Race & Ethnicity Bias Audit

## Selection Rate Results {-}

```{r race_table}
race_audit %>%
  mutate(
    selection_rate = percent(selection_rate, accuracy = 0.1),
    impact_ratio   = round(impact_ratio, 4)
  ) %>%
  select(-max_rate) %>%
  rename(
    "Race / Ethnicity"  = race_ethnicity,
    "Total Candidates"  = total_candidates,
    "Selected"          = selected,
    "Selection Rate"    = selection_rate,
    "Impact Ratio"      = impact_ratio,
    "LL144 Status"      = status
  ) %>%
  kable(caption = "Table 4: Race/Ethnicity Selection Rates & Impact Ratios") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width        = FALSE
  ) %>%
  row_spec(
    which(race_audit$status == "FAIL (Potential Bias)"),
    background = "#ffe0e0", bold = TRUE
  ) %>%
  row_spec(
    which(race_audit$status == "PASS"),
    background = "#e8f8f0"
  )
```

## Impact Ratio Chart {-}

```{r race_plot, fig.cap="Figure 2: Race/Ethnicity Impact Ratios — red dashed line marks the 80% legal threshold"}
ggplot(race_audit, aes(
    x    = reorder(race_ethnicity, -impact_ratio),
    y    = impact_ratio,
    fill = status
  )) +
  geom_col(width = 0.6, color = "white") +
  geom_hline(yintercept = 0.8, linetype = "dashed",
             color = "red", linewidth = 1) +
  annotate("text", x = 0.55, y = 0.84,
           label = "80% Legal Threshold",
           color = "red", size = 3.5, hjust = 0) +
  geom_text(
    aes(label = percent(impact_ratio, accuracy = 0.1)),
    vjust = -0.5, fontface = "bold", size = 4
  ) +
  scale_fill_manual(
    values = c("PASS"                  = "#A8E6CF",
               "FAIL (Potential Bias)" = "#FF8B94"),
    labels = c("PASS"                  = "✅ PASS",
               "FAIL (Potential Bias)" = "⚠️ FAIL")
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1.18)
  ) +
  labs(
    title    = "FairHire AI: Race/Ethnicity Impact Ratio Audit",
    subtitle = "NYC Local Law 144 Statutory Compliance Check",
    x        = "Demographic Group",
    y        = "Impact Ratio",
    fill     = "Compliance Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    legend.position = "top"
  )
```

---

# Bias Mitigation: Reweighing

When a group fails the 80% rule, NYC LL144 requires the employer to implement
a **bias mitigation strategy**. This section applies the **Reweighing**
pre-processing technique — the same approach used in IBM's AIF360 toolkit —
which adjusts candidate weights so underrepresented groups receive equitable
consideration without discarding any applications.

## Before vs. After Comparison Table {-}

```{r debiasing_table}
comparison %>%
  mutate(
    selection_rate = percent(selection_rate, accuracy = 0.1),
    impact_ratio   = round(impact_ratio, 4)
  ) %>%
  select(model, race_ethnicity, selected,
         selection_rate, impact_ratio, status) %>%
  rename(
    "Model Stage"      = model,
    "Race / Ethnicity" = race_ethnicity,
    "Selected"         = selected,
    "Selection Rate"   = selection_rate,
    "Impact Ratio"     = impact_ratio,
    "LL144 Status"     = status
  ) %>%
  kable(caption = "Table 5: Reweighing Debiasing — Before vs. After") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) %>%
  row_spec(
    which(comparison$status == "FAIL (Potential Bias)"),
    background = "#ffe0e0", bold = TRUE
  ) %>%
  row_spec(
    which(comparison$status == "PASS"),
    background = "#e8f8f0"
  ) %>%
  pack_rows("Before Debiasing", 1, nrow(race_audit_before)) %>%
  pack_rows("After Debiasing",
            nrow(race_audit_before) + 1, nrow(comparison))
```

## Before vs. After Chart {-}

```{r debiasing_plot, fig.cap="Figure 3: Impact Ratios before and after Reweighing bias mitigation", fig.height=6}
ggplot(comparison, aes(
    x    = reorder(race_ethnicity, -impact_ratio),
    y    = impact_ratio,
    fill = status
  )) +
  geom_col(width = 0.6, color = "white") +
  facet_wrap(~ model) +
  geom_hline(yintercept = 0.8, linetype = "dashed",
             color = "red", linewidth = 1) +
  annotate("text", x = 0.55, y = 0.84,
           label = "80% Threshold",
           color = "red", size = 3, hjust = 0) +
  geom_text(
    aes(label = percent(impact_ratio, accuracy = 0.1)),
    vjust = -0.5, fontface = "bold", size = 3.5
  ) +
  scale_fill_manual(
    values = c("PASS"                  = "#A8E6CF",
               "FAIL (Potential Bias)" = "#FF8B94"),
    labels = c("PASS"                  = "✅ PASS",
               "FAIL (Potential Bias)" = "⚠️ FAIL")
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1.18)
  ) +
  labs(
    title    = "FairHire AI: Bias Mitigation via Reweighing",
    subtitle = "Pre-processing debiasing — NYC Local Law 144 Compliance",
    x        = "Demographic Group",
    y        = "Impact Ratio",
    fill     = "Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    strip.text    = element_text(face = "bold", size = 11),
    legend.position = "top"
  )
```

---

# Compliance Summary & Recommendations

```{r compliance_summary}
# Final pass/fail summary table
bind_rows(
  gender_audit %>%
    mutate(protected_class = "Gender") %>%
    rename(group = gender),
  race_audit %>%
    mutate(protected_class = "Race / Ethnicity") %>%
    rename(group = race_ethnicity)
) %>%
  select(protected_class, group, selection_rate, impact_ratio, status) %>%
  mutate(
    selection_rate = percent(selection_rate, accuracy = 0.1),
    impact_ratio   = round(impact_ratio, 4)
  ) %>%
  rename(
    "Protected Class"  = protected_class,
    "Group"            = group,
    "Selection Rate"   = selection_rate,
    "Impact Ratio"     = impact_ratio,
    "LL144 Status"     = status
  ) %>%
  kable(caption = "Table 6: Full Compliance Summary — All Protected Classes") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width        = FALSE
  ) %>%
  row_spec(
    which(
      bind_rows(
        gender_audit %>% rename(group = gender),
        race_audit   %>% rename(group = race_ethnicity)
      )$status == "FAIL (Potential Bias)"
    ),
    background = "#ffe0e0", bold = TRUE
  )
```

**Recommendations:**

1. **Any group flagged ⚠️ FAIL** must be addressed before deployment under NYC LL144.
   Apply Reweighing (demonstrated in Section 4) or an in-processing fairness
   constraint to bring all impact ratios above 0.80.

2. **Re-audit annually** — NYC LL144 mandates a new third-party audit every
   calendar year and whenever the model is materially updated.

3. **Publish the audit summary** — results must be made publicly available on the
   employer's website at least 10 days before the tool is used in hiring decisions.

4. **Document mitigation actions** — retain records of all bias findings and
   remediation steps taken for regulatory review.

---
